---
title: "Step 3: Primer Trimming with Cutadapt"
output:
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Execute the pipeline once for HTML. The GitHub document is a code/reference
# view; disabling evaluation there avoids rerunning external tools and trying
# to screenshot HTML widgets in a non-HTML format.
render_target <- knitr::opts_knit$get("rmarkdown.pandoc.to")
knitr::opts_chunk$set(
  eval = !(grepl("^gfm", render_target) || grepl("markdown_github", render_target, fixed = TRUE)),
  #eval = FALSE,
  echo = TRUE
)
```

```{css custom-styles, echo=FALSE, eval=knitr::is_html_output()}
/* Custom styling for improved readability */
body {
  font-size: 16px;
  line-height: 1.6;
}

h1, h2, h3 {
  color: #2c3e50;
  margin-top: 1.5em;
}

code {
  background-color: #F2F2F2;
  color: #2c3e50;
  padding: 2px 6px;
  border-radius: 3px;
  font-size: 14px;
}

pre, pre code {
  font-size: 14px;
}

.alert-info {
  background-color: #2c3e50;
  color: #ffffff;
  border-left: 4px solid #f39c12;
  padding: 12px;
  margin: 15px 0;
}

.alert-warning {
  background-color: #2c3e50;
  color: #ffffff;
  border-left: 4px solid #e74c3c;
  padding: 12px;
  margin: 15px 0;
}

.alert-success {
  background-color: #d4edda;
  color: #155724;
  border-left: 4px solid #28a745;
  padding: 12px;
  margin: 15px 0;
}

/* Table cells wrap long content (long inline code, sentences) instead of
   overflowing or forcing horizontal scroll. */
table {
  width: 100%;
}

th, td {
  white-space: normal;
  word-wrap: break-word;
  overflow-wrap: break-word;
}

/* Style for primer sequence display */
.primer-seq {
  font-family: monospace;
  background-color: #f0f0f0;
  padding: 2px 6px;
  border-radius: 3px;
  font-size: 14px;
}
```

# Introduction

## Purpose

This notebook is **Step 3** of the sequencing data processing pipeline. It removes primer sequences from paired-end reads using [Cutadapt](https://cutadapt.readthedocs.io/), a widely-used tool for finding and removing adapter sequences, primers, and other unwanted sequences from high-throughput sequencing reads. It produces a detailed, auditable account of primer content before and after trimming. Length filtering, N-filtering, and quality filtering are handled downstream in the [DADA2 pipeline (Step 5)](5_dada2_pipeline.md).

Primer sequences must be removed before downstream analysis because:

- **Primers are not biological sequences**: They are synthetic oligonucleotides added during PCR
- **They introduce bias**: Primers can inflate sequence similarity scores
- **[DADA2](https://benjjneb.github.io/dada2/) requires primer-free reads**: its error model works best on amplicon sequences only
- **Taxonomic assignment accuracy**: Primers can interfere with database matching

## Prerequisites

Before running this notebook, ensure that:

1.  FASTQ files (`.fastq`, `.fq`, and their gzip-compressed variants, case-insensitively) are present in the [data/fastq/](../../data/fastq/) directory
2.  **Cutadapt** is installed in the [tools/cutadapt/](../../tools/cutadapt/) R project directory by running the [setup/install_required_tools.R](../../setup/install_required_tools.R) script
3.  **[Step 1 (Data Integrity Check)](1_data_integrity_check.md)** has been completed successfully (**optional**)
4.  **[Step 2 (FastQC Quality Reports)](2_fastqc_quality_reports.md)** has been completed successfully (**optional**)

## What This Notebook Does

1.  **Discovers FASTQ Files**: Scans the input directory for paired-end FASTQ files
2.  **Counts Primers Before Trimming**: Uses a custom, anchored search to count, for every sample, primer, read file, and sequence orientation, how many reads carry a primer at a read end
3.  **Runs Cutadapt**: Trims primer sequences from both forward and reverse reads, writing the primer-trimmed FASTQ files
4.  **Counts Primers After Trimming**: Repeats the anchored count on the trimmed reads, so before/after and the number removed are directly comparable per sample
5.  **Exports Results**: Saves the primer-hit table to an Excel workbook

This notebook does **primer trimming only**. Length filtering (removing reads shorter than 100 bp), N-filtering, and quality filtering are all handled downstream in the [DADA2 pipeline (Step 5)](5_dada2_pipeline.md), where the read losses at each stage are tracked together.

## Expected Input

- **Location**: FASTQ files should be placed in the [data/fastq/](../../data/fastq/) directory relative to your [project root](../../)
- **Naming Convention**: Paired files must follow a consistent naming pattern, for example:
  - Forward reads: `{sample_id}_L001_R1_001.fastq.gz` or `{sample_id}_L001_R1_001.fq`
  - Reverse reads: `{sample_id}_L001_R2_001.fastq.gz` or `{sample_id}_L001_R2_001.fq`
  - The pipeline is intentionally hardcoded to lane token `L001`. `.fastq`, `.fq`, and their gzip-compressed variants are detected case-insensitively.
  - This workflow expects one R1/R2 pair per `SampleID`. Merge lane-split files and name the merged pair with the `L001` token before running the notebook.
- **Format**: Standard FASTQ format (4-line records per read)

## Expected Output

- Primer-trimmed reads in [results/3_cutadapt_primer_trimming/primer_trimmed_reads/](../../results/3_cutadapt_primer_trimming/primer_trimmed_reads/) (input to the [Step 4](4_dada2_parameter_selection.md) and [Step 5](5_dada2_pipeline.md))
- Cutadapt log files for each sample in [results/3_cutadapt_primer_trimming/cutadapt_logs/](../../results/3_cutadapt_primer_trimming/cutadapt_logs/)
- Summary Excel file ([primer_trimming_summary.xlsx](../../results/3_cutadapt_primer_trimming/primer_trimming_summary.xlsx)) with per-sample primer-hit counts before and after trimming

::: {.alert .alert-warning}
**Downstream note**: The FASTQ files in [primer_trimmed_reads/](../../results/3_cutadapt_primer_trimming/primer_trimmed_reads/) are primer-trimmed only. Length filtering (<100 bp), N-filtering, and quality filtering are performed in the [DADA2 pipeline (Step 5)](5_dada2_pipeline.md), which takes this folder as its input.
:::

------------------------------------------------------------------------

# Environment Setup

## Load Required Packages {#load-packages}

The chunk below loads every R package this notebook depends on, plus three helper functions shared across the whole pipeline: Excel writing, column-dictionary documentation, and clickable output links (each sourced from [R/functions/](../functions/)).

```{r load-packages, message=FALSE, warning=FALSE}
# Biostrings: Biological string manipulation (Bioconductor)
# Provides functions for handling DNA/RNA sequences, generating orientations,
# and pattern counting (vcountPattern) for the custom primer-hit counter
library(Biostrings)

# data.table: High-performance data manipulation
# Provides fast operations for handling large result tables
library(data.table)

# DT: Interactive tables in R Markdown
# Creates searchable, sortable HTML tables
library(DT)

# fs: Cross-platform filesystem operations
# Consistent functions for file and directory manipulation
library(fs)

# here: Project-relative file paths
# Enables reproducible path construction regardless of working directory
library(here)

# openxlsx: Excel file creation and manipulation
# Read and write Excel files without Java dependencies
library(openxlsx)

# parallel: Parallel processing support
# Built-in R package for multi-core processing
library(parallel)

# ShortRead: FASTQ file handling (Bioconductor)
# Efficient tools for reading and counting FASTQ files
library(ShortRead)

# stringr: Consistent string manipulation
# Intuitive string processing functions
library(stringr)

# Source our custom Excel utility function from the project's function library
# This function handles creating or appending sheets to Excel workbooks
source(here("R", "functions", "add_sheet_to_excel_function.R"))

# Source our custom column-dictionary builder from the project's function library
# This function documents every column of a sheet being exported to Excel
source(here("R", "functions", "build_column_dictionary_function.R"))

# Source our custom output-links helper from the project's function library
# This function renders clickable Markdown links to this notebook's output files
source(here("R", "functions", "render_output_links_function.R"))

# Source the custom render_output_tree utility function from the project's
# function library. This function prints this notebook's entire output
# folder as a clickable directory tree (used in "Output File Summary" below),
# scanned live from disk at knit time so it always matches what was actually
# produced on this run.
source(here("R", "functions", "render_output_tree_function.R"))
```

------------------------------------------------------------------------

# Configuration

## Define Path Parameters {#define-paths}

Set up the input and output directory paths. Using `here()` ensures paths are relative to the [project root](../../), making the script portable across different systems.

```{r define-paths}
# Input folder containing raw FASTQ files
# This should be the same directory used in Steps 1 and 2
fastq_input_folder <- here("data", "fastq")

# Base results folder for all pipeline outputs
results_folder <- here("results")

# Specific output folder for this step (Step 3: Primer Trimming)
output_folder <- here(results_folder, "3_cutadapt_primer_trimming")

# Output folder for the primer-trimmed reads. This is the final output of this
# step and the input the DADA2 pipeline (Step 5) consumes.
primer_trimmed_folder <- here(output_folder, "primer_trimmed_reads")

# Folder for cutadapt log files (the human-readable record of what cutadapt did)
log_folder <- here(output_folder, "cutadapt_logs")

# Output Excel filename for primer trimming statistics
output_excel_filename <- "primer_trimming_summary.xlsx"

# Construct the full path to the output Excel file
output_excel_path <- here(output_folder, output_excel_filename)
```

## Define File Pattern Parameters {#define-patterns}

Configure the file naming patterns used to identify forward and reverse reads.

```{r define-patterns}
# Token identifying forward (Read 1) files in paired-end sequencing,
# written as a regular expression (not a literal string) so it matches
# .fastq/.fq and optional gzip compression for the required L001 lane token
# automatically -- see data/README.md's "Naming Convention" section.
# (?i) makes the extension match case-insensitively, and $ anchors the match
# to the end of the filename.
forward_pattern <- "(?i)_L001_R1_001\\.(fastq|fq)(\\.gz)?$"

# Token identifying reverse (Read 2) files in paired-end sequencing -- see
# forward_pattern above for why this is a regex rather than a literal
# extension.
reverse_pattern <- "(?i)_L001_R2_001\\.(fastq|fq)(\\.gz)?$"

# Display configuration
cat("File Pattern Configuration:\n",
    "- Forward reads pattern:", forward_pattern, "\n",
    "- Reverse reads pattern:", reverse_pattern, "\n")
```

## Define Primer Sequences {#define-primers}

Specify the primer sequences used for 16S amplicon generation. These primers will be trimmed from the reads.

::: {.alert .alert-warning}
**Important**: Update these sequences to match the primers used in your experiment. The default primers below target the V3-V4 region of the 16S rRNA gene.
:::

```{r define-primers}
# Forward primer sequence (5' -> 3') targeting the desired 16S variable region
# IUPAC ambiguity codes: N = any, W = A/T
forward_primer <- "CCTACGGGNGGCWGCAG"

# Reverse primer sequence (5' -> 3') targeting the desired 16S variable region
# IUPAC ambiguity codes: H = A/C/T, V = A/C/G
reverse_primer <- "GACTACHVGGGTATCTAATCC"

# Reverse complements of each primer. These are the sequences that appear at
# the 3' END of a read when the insert is shorter than the read length and the
# sequencer reads through into the opposite primer (read-through). They are
# used by cutadapt (-a / -A) to trim that read-through.
fwd_primer_rc <- as.character(reverseComplement(DNAString(forward_primer)))
rev_primer_rc <- as.character(reverseComplement(DNAString(reverse_primer)))

# Display primer information
cat("Primer Configuration:\n",
    "- Forward primer (5'->3'):", forward_primer, "\n",
    "- Reverse primer (5'->3'):", reverse_primer, "\n",
    "- Forward primer length:", nchar(forward_primer), "bp\n",
    "- Reverse primer length:", nchar(reverse_primer), "bp\n",
    "- Forward primer reverse complement:", fwd_primer_rc, "\n",
    "- Reverse primer reverse complement:", rev_primer_rc, "\n")
```

## Define Processing Parameters {#define-params}

Configure parameters for parallel processing, primer counting, and cutadapt execution.

```{r define-processing-params}
# Determine the number of CPU threads for parallel processing
# Reserve 2 cores for system operations to maintain responsiveness
detected_cores <- suppressWarnings(detectCores())
available_cores <- if (length(detected_cores) == 1L && is.finite(detected_cores) && detected_cores >= 1) as.integer(detected_cores) else 1L
nr_threads <- max(1L, available_cores - 2L)

# Anchor buffer (bp) for the custom primer-hit counter. A read counts as a hit
# only if a primer orientation is found within the first or last
# (primer_length + primer_anchor_buffer) bases. Kept deliberately small (2 bp)
# so the window is precise -- just wide enough to tolerate a minor positional
# offset (e.g. a stray leading base in some library preps) without letting the
# search drift inward and pick up coincidental interior matches of the
# degenerate primer. Raise to a slightly larger value only if primers are known
# to sit several bases from the read end.
primer_anchor_buffer <- 2L

# Number of reads loaded from each FASTQ at a time while counting primer hits.
# This bounds memory use independently of total file size.
fastq_chunk_size <- 100000L

# Path to the project-local Cutadapt executable
cutadapt_path <- here("tools", "cutadapt", "venv", "bin", "cutadapt")

if (!dir_exists(fastq_input_folder)) {
  stop("Raw FASTQ input folder not found: ", fastq_input_folder,
       "\nComplete Step 1 and verify the project paths before running Step 3.")
}
if (!file_exists(cutadapt_path) || file.access(cutadapt_path, mode = 1) != 0) {
  stop("Cutadapt executable not found or not executable: ", cutadapt_path,
       "\nRun setup/install_required_tools.R.")
}

# Verify that the executable can actually start. Merely checking the executable
# bit is insufficient for a copied Python virtual environment because its
# launcher can retain an absolute interpreter path from another machine.
cutadapt_version_log <- tempfile("cutadapt-version-", fileext = ".log")
cutadapt_version_status <- system2(
  cutadapt_path,
  args = "--version",
  stdout = cutadapt_version_log,
  stderr = cutadapt_version_log
)
cutadapt_version <- readLines(cutadapt_version_log, warn = FALSE)
unlink(cutadapt_version_log)
if (!identical(cutadapt_version_status, 0L)) {
  stop("Cutadapt could not be started (exit status ", cutadapt_version_status, "): ",
       paste(cutadapt_version, collapse = "\n"),
       "\nRecreate the project-local tools with setup/install_required_tools.R --force-reinstall.")
}

# Display configuration
cat("Processing Configuration:\n",
    "- Available CPU cores:", available_cores, "\n",
    "- Threads for processing:", nr_threads, "\n",
    "- Primer anchor buffer (bp):", primer_anchor_buffer, "\n",
    "- FASTQ counting chunk size:", fastq_chunk_size, "reads\n",
    "- Cutadapt path:", cutadapt_path, "\n")
```

------------------------------------------------------------------------

# Initialize Output Directories

Create the output folder structure for primer trimming results.

```{r create-output-dirs}
# Create output directories if they don't exist
# Using fs::dir_create which is safe to run multiple times

# Main results folder
if (!dir_exists(results_folder)) {
  dir_create(results_folder, recurse = TRUE)
  cat("Created results directory:", results_folder, "\n")
}

# Step 3 output folder
if (!dir_exists(output_folder)) {
  dir_create(output_folder, recurse = TRUE)
  cat("Created output directory:", output_folder, "\n")
} else {
  cat("Using existing output directory:", output_folder, "\n")
}

```

------------------------------------------------------------------------

# Define Helper Functions

## Sample ID Extraction {#func-sample-id}

This function extracts the sample identifier from a FASTQ filename.

```{r func-extract-sample-id}
# Extract Sample ID from FASTQ Filename
#
# Parses a FASTQ filename and returns the sample identifier portion.
# Assumes sample ID is the first component before any underscore.
#
# Arguments:
#   filename - Character string of the filename (without path)
#   pattern  - The file pattern to remove, as a regular expression (e.g.
#              forward_pattern/reverse_pattern above, which match both
#              compressed and uncompressed FASTQ extensions)
# Returns:
#   Character string containing the sample ID
#
# Example:
#   extract_sample_id("Sample001_L001_R1_001.fq.gz", forward_pattern)
#   # Returns "Sample001"
extract_sample_id <- function(filename, pattern) {
  # Remove the pattern from the filename. `pattern` is matched as a
  # regular expression here (NOT fixed()/literal matching) since
  # forward_pattern/reverse_pattern above are regexes with an optional
  # gzip-suffix group.
  sample_name <- str_remove(filename, pattern)
  # Extract first part before underscore (if present)
  sample_id <- str_split(sample_name, "_", simplify = TRUE)[1]
  return(sample_id)
}
```

## Primer Orientation Generator {#func-primer-orientations}

This function generates all four orientations of a primer sequence (Forward, Complement, Reverse, ReverseComplement). Counting all four is the standard DADA2 sanity check: only two of them have a real biological mechanism (Forward = the primer itself at the 5' end; ReverseComplement = read-through at the 3' end), while non-zero counts in the Complement/Reverse columns would signal an unexpected library-prep or contamination issue.

```{r func-primer-orientations}
# Generate All Primer Orientations
#
# Creates all four possible orientations of a DNA primer sequence, returned as
# a named character vector so downstream tables can label each orientation.
#
# Arguments:
#   primer - Character string of the primer sequence (5' -> 3')
# Returns:
#   Named character vector: Forward, Complement, Reverse, ReverseComplement
get_primer_orientations <- function(primer) {
  dna_seq <- DNAString(primer)
  c(
    Forward           = as.character(dna_seq),
    Complement        = as.character(complement(dna_seq)),
    Reverse           = as.character(reverse(dna_seq)),
    ReverseComplement = as.character(reverseComplement(dna_seq))
  )
}
```

## Anchored Primer-Hit Counter {#func-primer-hits}

This function counts how many reads carry a given primer orientation at a read **end**, honoring IUPAC ambiguity codes but requiring an exact (zero-mismatch) match.

::: {.alert .alert-info}
**Note on the Custom Primer Counter (anchored matching)**: The before/after primer counts in this notebook are produced by a custom R search built on `Biostrings::vcountPattern()`.

- The search is **anchored to the read ends**: a read only counts as a hit if the primer (or one of its orientations) is found within the first or last `primer_length + primer_anchor_buffer` bases, which is where primers and read-through actually occur.

- A small buffer (default **2 bp**) allows for minor positional offsets -- for example a stray leading base before the primer -- without opening the search up to the coincidental *interior* matches of a short, IUPAC-degenerate primer that made earlier whole-read counts fail to drop to zero after trimming.

- Matching is **exact** (IUPAC-aware but zero-mismatch), so a read whose primer region carries a sequencing error may not be counted; these counts are therefore a conservative lower bound.

- The cutadapt log files (`cutadapt_logs/`) hold cutadapt's own error-tolerant tally if needed.
:::

```{r func-primer-hits}
# Count Reads with a Primer Orientation at Either Read End
#
# For a single already-loaded set of reads (a DNAStringSet), counts how many
# reads contain `orientation_seq` within the first or last
# (nchar(orientation_seq) + anchor_buffer) bases. IUPAC codes in the query are
# honored (fixed = FALSE); matching is exact (no mismatches), so counts are a
# conservative lower bound relative to cutadapt's error-tolerant matching.
#
# Arguments:
#   reads           - DNAStringSet from one streamed FASTQ chunk
#   read_widths     - Integer vector of read lengths (width(reads)); passed in
#                      so it is computed once per file rather than once per
#                      orientation
#   orientation_seq - Character string: the primer orientation to search for
#   anchor_buffer   - Integer buffer (bp) added to the primer length to define
#                      the 5' and 3' search windows
# Returns:
#   Integer count of reads with at least one anchored hit
count_anchored_orientation_hits <- function(reads, read_widths, orientation_seq, anchor_buffer) {
  primer_len  <- nchar(orientation_seq)
  window_len  <- primer_len + anchor_buffer

  # 5' window: bases 1 .. min(window_len, read_width)
  five_prime_window <- subseq(reads,
                              start = 1L,
                              end   = pmin(window_len, read_widths))

  # 3' window: last window_len bases, floored at base 1
  three_prime_window <- subseq(reads,
                               start = pmax(1L, read_widths - window_len + 1L),
                               end   = read_widths)

  # A read is a hit if the orientation is found in either anchored window
  hits_5 <- vcountPattern(orientation_seq, five_prime_window, fixed = FALSE) > 0
  hits_3 <- vcountPattern(orientation_seq, three_prime_window, fixed = FALSE) > 0
  sum(hits_5 | hits_3)
}

# Count Primer Hits in One FASTQ File (long format)
#
# Streams bounded chunks rather than loading the complete file into memory.
count_fastq_primer_hits <- function(fastq_path, primer_specs, read_label,
                                    sample_id, anchor_buffer, chunk_size) {
  hit_totals <- setNames(
    integer(sum(lengths(primer_specs))),
    unlist(lapply(names(primer_specs), function(primer_label) {
      paste(primer_label, names(primer_specs[[primer_label]]), sep = "::")
    }), use.names = FALSE)
  )

  streamer <- FastqStreamer(fastq_path, n = chunk_size)
  on.exit(close(streamer), add = TRUE)

  repeat {
    fastq_chunk <- yield(streamer)
    if (length(fastq_chunk) == 0L) break
    reads <- sread(fastq_chunk)
    read_widths <- width(reads)

    for (primer_label in names(primer_specs)) {
      for (orientation_label in names(primer_specs[[primer_label]])) {
        key <- paste(primer_label, orientation_label, sep = "::")
        hit_totals[[key]] <- hit_totals[[key]] + count_anchored_orientation_hits(
          reads, read_widths, primer_specs[[primer_label]][[orientation_label]], anchor_buffer
        )
      }
    }
  }

  rbindlist(lapply(names(hit_totals), function(key) {
    key_parts <- str_split_fixed(key, fixed("::"), 2L)
    data.table(
      SampleID = sample_id,
      Primer = key_parts[1L],
      Read_File = read_label,
      Orientation = key_parts[2L],
      Hits = unname(hit_totals[[key]])
    )
  }))
}

# Count Primer Hits Across All Samples (long format)
#
# Builds a tidy per-sample / per-primer / per-read-file / per-orientation table
# of anchored primer-hit counts. Each FASTQ file is read exactly once; all
# primer/orientation counts for that file reuse the loaded reads.
#
# Arguments:
#   fwd_paths, rev_paths           - Character vectors of R1/R2 FASTQ paths
#                                     (paired)
#   sample_ids                     - Character vector of sample identifiers
#                                     (paired with paths)
#   fwd_orientations, rev_orientations - Named character vectors from
#                                     get_primer_orientations() for the
#                                     forward and reverse primer
#   anchor_buffer                  - Integer buffer passed to the anchored
#                                     counter
#   n_threads                      - Number of cores for mclapply
#   chunk_size                     - Maximum reads held per FASTQ stream chunk
# Returns:
#   data.table with columns SampleID, Primer, Read_File, Orientation, Hits
count_primer_hits_long <- function(fwd_paths, rev_paths, sample_ids,
                                   fwd_orientations, rev_orientations,
                                   anchor_buffer, n_threads, chunk_size) {
  per_sample <- mclapply(seq_along(sample_ids), function(i) {
    primer_specs <- list(FWD = fwd_orientations, REV = rev_orientations)
    rbindlist(list(
      count_fastq_primer_hits(fwd_paths[i], primer_specs, "R1", sample_ids[i],
                              anchor_buffer, chunk_size),
      count_fastq_primer_hits(rev_paths[i], primer_specs, "R2", sample_ids[i],
                              anchor_buffer, chunk_size)
    ))
  }, mc.cores = if (.Platform$OS.type == "windows") 1L else n_threads)

  rbindlist(per_sample)
}
```

------------------------------------------------------------------------

# Discover FASTQ Files

## List Input Files {#list-files}

Scan the input directory for paired-end FASTQ files.

```{r list-fastq-files}
# Get all forward read file paths
fwd_raw_paths <- dir_ls(
  path = fastq_input_folder,
  regexp = forward_pattern
) %>%
  # Sort for consistent ordering
  sort()

# Get all reverse read file paths
rev_raw_paths <- dir_ls(
  path = fastq_input_folder,
  regexp = reverse_pattern
) %>%
  sort()

# Extract sample names from forward files
sample_names <- sapply(
  basename(fwd_raw_paths),
  extract_sample_id,
  pattern = forward_pattern,
  USE.NAMES = FALSE
)

# Display discovery results
cat("File Discovery Results:\n",
    "- Input directory:", fastq_input_folder, "\n",
    "- Forward files found:", length(fwd_raw_paths), "\n",
    "- Reverse files found:", length(rev_raw_paths), "\n",
    "- Samples identified:", length(sample_names), "\n")
```

## Validate File Pairing {#validate-pairing}

Verify that all samples have both forward and reverse read files.

```{r validate-file-pairing}
# Compare complete filename stems first; comparing only the text before the
# first underscore can conceal a mismatched pair with the same SampleID.
fwd_pair_stems <- str_remove(basename(fwd_raw_paths), forward_pattern)
rev_pair_stems <- str_remove(basename(rev_raw_paths), reverse_pattern)

# Extract sample names from reverse files for comparison
rev_sample_names <- sapply(
  basename(rev_raw_paths),
  extract_sample_id,
  pattern = reverse_pattern,
  USE.NAMES = FALSE
)

# Check that a non-empty set of complete forward/reverse stems matches.
if (length(fwd_raw_paths) == 0L || length(rev_raw_paths) == 0L) {
  stop("No paired FASTQ files matched the configured patterns in: ", fastq_input_folder,
       "\nForward pattern: ", forward_pattern,
       "\nReverse pattern: ", reverse_pattern)
}
if (length(fwd_pair_stems) != length(rev_pair_stems) ||
    !identical(fwd_pair_stems, rev_pair_stems)) {
  stop(
    "Mismatch between forward and reverse files!\n",
    "Forward samples: ", length(sample_names), "\n",
    "Reverse samples: ", length(rev_sample_names), "\n",
    "Forward-only stems: ", paste(setdiff(fwd_pair_stems, rev_pair_stems), collapse = ", "), "\n",
    "Reverse-only stems: ", paste(setdiff(rev_pair_stems, fwd_pair_stems), collapse = ", "), "\n",
    "Please check your input files."
  )
}
if (anyDuplicated(sample_names)) {
  stop("Sample IDs are not unique before the first underscore: ",
       paste(unique(sample_names[duplicated(sample_names)]), collapse = ", "))
}

# Create sample inventory table
sample_inventory <- data.table(
  SampleID = sample_names,
  Forward_File = basename(fwd_raw_paths),
  Reverse_File = basename(rev_raw_paths)
)

# Display sample inventory
cat("Sample Inventory:\n",
    "- Total samples:", length(sample_names), "\n")
```

```{r echo=FALSE}
datatable(sample_inventory,
          options = list(pageLength = 10, scrollX = TRUE,
                          scrollY = "400px", scrollCollapse = TRUE, paging = FALSE),
          caption = "Input FASTQ files for primer trimming")
```

------------------------------------------------------------------------

# Generate Primer Orientations

Create all four orientations of the forward and reverse primers, which the anchored counter searches for in each read file.

```{r generate-primer-orientations}
fwd_primer_orientations <- get_primer_orientations(forward_primer)
rev_primer_orientations <- get_primer_orientations(reverse_primer)

primer_orientation_table <- data.table(
  Orientation    = names(fwd_primer_orientations),
  Forward_Primer = unname(fwd_primer_orientations),
  Reverse_Primer = unname(rev_primer_orientations)
)

cat("Primer orientations generated (searched for at read ends):\n\n")
```

```{r echo=FALSE}
datatable(primer_orientation_table,
          options = list(dom = 't', scrollY = "400px", scrollCollapse = TRUE, paging = FALSE),
          rownames = FALSE,
          caption = "All four orientations of each primer")
```

------------------------------------------------------------------------

# Count Primers Before Trimming {#count-before}

Count primer occurrences at the read ends of the **raw** reads, for every sample, primer, read file, and orientation. This establishes the baseline that the after-trimming counts are compared against.

::: {.alert .alert-info}
**Note**: This step reads every raw FASTQ file once. It may take several minutes for large datasets.
:::

```{r echo=FALSE}
cat("Counting primer occurrences before trimming (anchored search)...\n\n")
```

```{r count-primers-before, message=FALSE}
primer_hits_before_dt <- count_primer_hits_long(
  fwd_paths        = fwd_raw_paths,
  rev_paths        = rev_raw_paths,
  sample_ids       = sample_names,
  fwd_orientations = fwd_primer_orientations,
  rev_orientations = rev_primer_orientations,
  anchor_buffer    = primer_anchor_buffer,
  n_threads        = nr_threads,
  chunk_size       = fastq_chunk_size
)

cat("Pre-trimming primer counting complete (",
    nrow(primer_hits_before_dt), "rows ).\n")
```

------------------------------------------------------------------------

# Run Cutadapt {#run-cutadapt}

## Verify Cutadapt Installation {#verify-cutadapt}

Check that Cutadapt is available and display its version.

```{r verify-cutadapt}
cat("Cutadapt Configuration:\n",
    "- Path:", cutadapt_path, "\n",
    "- Version:", cutadapt_version, "\n")
```

## Execute Primer Trimming

Run Cutadapt on each sample to remove primer sequences from paired-end reads. The primer-trimmed reads are written to the [primer_trimmed_reads/](../../results/3_cutadapt_primer_trimming/primer_trimmed_reads/) directory, which the [DADA2 pipeline (Step 5)](5_dada2_pipeline.md) consumes.

::: {.alert .alert-info}
**Note**: Cutadapt is run with the following settings:

- `-g PRIMER`: Trim 5' forward primer from forward reads
- `-a PRIMER_RC`: Trim 3' read-through (reverse-complement of reverse primer) from forward reads
- `-G PRIMER`: Trim 5' reverse primer from reverse reads
- `-A PRIMER_RC`: Trim 3' read-through (reverse-complement of forward primer) from reverse reads
- `-n 2`: Allow up to 2 primer sequences to be removed per read
- `--cores N`: Use the configured `nr_threads` limit while reserving two cores when available
:::

```{r run-cutadapt, message=FALSE}
# Write this run into an isolated staging tree. The final folders consumed by
# downstream steps are replaced only after every Cutadapt command succeeds.
staging_folder <- path(
  output_folder,
  paste0(".cutadapt-staging-", format(Sys.time(), "%Y%m%d%H%M%S"), "-", Sys.getpid())
)
staging_trimmed_folder <- path(staging_folder, "primer_trimmed_reads")
staging_log_folder <- path(staging_folder, "cutadapt_logs")
dir_create(staging_trimmed_folder, recurse = TRUE)
dir_create(staging_log_folder, recurse = TRUE)

# Define output paths for cutadapt's primer-trimmed files
fwd_staging_paths <- path(staging_trimmed_folder, basename(fwd_raw_paths))
rev_staging_paths <- path(staging_trimmed_folder, basename(rev_raw_paths))

cat("Running Cutadapt for primer trimming...\n",
    "Processing", length(sample_names), "samples.\n\n")

# Run cutadapt for each sample
for (i in seq_along(sample_names)) {

  cat("Processing sample", i, "of", length(sample_names), ":", sample_names[i], "\n")

  # Define log file path
  log_file <- path(staging_log_folder, paste0(sample_names[i], "_cutadapt.log"))

  # Build cutadapt command arguments
  cutadapt_args <- c(
    # Forward read trimming:
    # -g: Trim forward primer from 5' end of forward reads
    # -a: Trim reverse complement of reverse primer from 3' end (read-through)
    "-g", shQuote(forward_primer),
    "-a", shQuote(rev_primer_rc),

    # Reverse read trimming:
    # -G: Trim reverse primer from 5' end of reverse reads
    # -A: Trim reverse complement of forward primer from 3' end (read-through)
    "-G", shQuote(reverse_primer),
    "-A", shQuote(fwd_primer_rc),

    # Allow up to 2 primers to be removed per read
    "-n", "2",

    # Respect the same bounded thread setting reported above.
    paste0("--cores=", nr_threads),

    # Output files
    "-o", shQuote(as.character(fwd_staging_paths[i])),
    "-p", shQuote(as.character(rev_staging_paths[i])),

    # Input files (raw reads)
    shQuote(as.character(fwd_raw_paths[i])),
    shQuote(as.character(rev_raw_paths[i]))
  )

  # Execute cutadapt and capture output
  result <- system2(
    cutadapt_path,
    args = cutadapt_args,
    stdout = log_file,
    stderr = log_file
  )

  # Check for errors
  if (result != 0 || !file_exists(fwd_staging_paths[i]) || !file_exists(rev_staging_paths[i]) ||
      file_size(fwd_staging_paths[i]) == 0 || file_size(rev_staging_paths[i]) == 0) {
    stop("Cutadapt failed for sample ", sample_names[i],
         " (exit code ", result, "). See log: ", log_file)
  }
}

# Publish the complete run as one dataset only after every sample succeeded.
# Preserve the previous run inside staging until both replacements succeed so a
# filesystem error during publication can roll back to the complete old run.
previous_trimmed_backup <- path(staging_folder, "previous_primer_trimmed_reads")
previous_log_backup <- path(staging_folder, "previous_cutadapt_logs")
tryCatch({
  if (dir_exists(primer_trimmed_folder)) file_move(primer_trimmed_folder, previous_trimmed_backup)
  if (dir_exists(log_folder)) file_move(log_folder, previous_log_backup)
  file_move(staging_trimmed_folder, primer_trimmed_folder)
  file_move(staging_log_folder, log_folder)
}, error = function(e) {
  if (dir_exists(primer_trimmed_folder)) dir_delete(primer_trimmed_folder)
  if (dir_exists(log_folder)) dir_delete(log_folder)
  if (dir_exists(previous_trimmed_backup)) file_move(previous_trimmed_backup, primer_trimmed_folder)
  if (dir_exists(previous_log_backup)) file_move(previous_log_backup, log_folder)
  stop("Could not publish the completed Cutadapt run; the previous output was restored: ",
       conditionMessage(e))
})
dir_delete(staging_folder)

fwd_trimmed_paths <- path(primer_trimmed_folder, basename(fwd_raw_paths))
rev_trimmed_paths <- path(primer_trimmed_folder, basename(rev_raw_paths))
```

```{r echo=FALSE}
cat("\nCutadapt primer trimming complete.\n",
    "Log files saved to:", log_folder, "\n")

render_output_links(log_folder, labels = "Cutadapt log files (folder)")
```

------------------------------------------------------------------------

# Count Primers After Trimming {#count-after}

Repeat the anchored count on the **cutadapt-trimmed** reads. Comparing these to the pre-trimming counts shows how many primer-bearing reads were cleaned per sample, primer, read file, and orientation.

```{r echo=FALSE}
cat("Counting primer occurrences after trimming (anchored search)...\n\n")
```

```{r count-primers-after, message=FALSE}
primer_hits_after_dt <- count_primer_hits_long(
  fwd_paths        = fwd_trimmed_paths,
  rev_paths        = rev_trimmed_paths,
  sample_ids       = sample_names,
  fwd_orientations = fwd_primer_orientations,
  rev_orientations = rev_primer_orientations,
  anchor_buffer    = primer_anchor_buffer,
  n_threads        = nr_threads,
  chunk_size       = fastq_chunk_size
)

cat("Post-trimming primer counting complete (",
    nrow(primer_hits_after_dt), "rows ).\n")
```

## Build Combined Before/After Primer-Hit Table {#primer-hit-table}

Merge the before and after counts into one tidy table, adding the number of primer-bearing reads removed for each combination.

```{r build-primer-hit-table}
# Merge on the four identifying keys; rename the Hits columns to Before/After
primer_hit_counts_dt <- merge(
  setnames(copy(primer_hits_before_dt), "Hits", "Hits_Before"),
  setnames(copy(primer_hits_after_dt),  "Hits", "Hits_After"),
  by = c("SampleID", "Primer", "Read_File", "Orientation")
)

# Reads that carried a primer before but not after
primer_hit_counts_dt[, Hits_Removed := Hits_Before - Hits_After]

# Percentage of primer-bearing reads removed, relative to the pre-trimming
# count for that same sample/primer/read-file/orientation. When Hits_Before is
# 0 there was nothing to remove, so the percentage is undefined (0/0) and is
# reported as NA rather than NaN.
primer_hit_counts_dt[, Pct_Removed := fifelse(
  Hits_Before > 0, round(100 * Hits_Removed / Hits_Before, 2), NA_real_
)]

# Order rows so each sample's block reads FWD-then-REV, R1-then-R2, in the
# canonical orientation order
primer_hit_counts_dt[, Orientation := factor(
  Orientation, levels = c("Forward", "Complement", "Reverse", "ReverseComplement")
)]
setorder(primer_hit_counts_dt, SampleID, Primer, Read_File, Orientation)
primer_hit_counts_dt[, Orientation := as.character(Orientation)]

# Console summary: totals for the four biologically meaningful combinations,
# summed across samples (see "Understanding the Output" for why these four)
meaningful_combos <- rbind(
  data.table(Primer = "FWD", Read_File = "R1", Orientation = "Forward"),
  data.table(Primer = "REV", Read_File = "R2", Orientation = "Forward"),
  data.table(Primer = "REV", Read_File = "R1", Orientation = "ReverseComplement"),
  data.table(Primer = "FWD", Read_File = "R2", Orientation = "ReverseComplement")
)
# Flag which rows are biologically expected (a real primer / read-through
# mechanism) versus not (no mechanism -- contaminant or other artefact).
# meaningful_combos (above) lists the four combinations with a real mechanism;
# every other row is a contaminant/artefact sanity-net cell.
primer_hit_counts_dt[, Biological_Relevance := "Contaminant / other"]
primer_hit_counts_dt[meaningful_combos, on = c("Primer", "Read_File", "Orientation"),
                     Biological_Relevance := "Biologically relevant"]
setcolorder(primer_hit_counts_dt,
            c("SampleID", "Primer", "Read_File", "Orientation", "Biological_Relevance"))

meaningful_totals <- merge(primer_hit_counts_dt, meaningful_combos,
                           by = c("Primer", "Read_File", "Orientation"))[
  , .(Hits_Before = sum(Hits_Before), Hits_After = sum(Hits_After)),
  by = .(Primer, Read_File, Orientation)]

cat("\nPRIMER HITS (anchored, exact match) -- biologically expected values:\n")
```

```{r echo=FALSE}
for (r in seq_len(nrow(meaningful_totals))) {
  cat(sprintf("  %s primer, %s, %-17s : before %s  ->  after %s\n",
              meaningful_totals$Primer[r], meaningful_totals$Read_File[r],
              meaningful_totals$Orientation[r],
              format(meaningful_totals$Hits_Before[r], big.mark = ","),
              format(meaningful_totals$Hits_After[r], big.mark = ",")))
}
```

```{r echo=FALSE}
datatable(primer_hit_counts_dt,
          options = list(pageLength = 16, scrollX = TRUE,
                          scrollY = "400px", scrollCollapse = TRUE, paging = FALSE),
          rownames = FALSE,
          caption = "Anchored primer-hit counts per sample, primer, read file, and orientation (before, after, removed, percent removed)")
```

------------------------------------------------------------------------

# Export Results

## Save to Excel {#save-excel}

Export all results to an Excel workbook with multiple sheets.

```{r export-to-excel}
# Recreate this run's workbook so sheets left by older notebook versions cannot
# survive and be mistaken for current output.
if (file_exists(output_excel_path)) file_delete(output_excel_path)

# Sheet 1: Custom anchored primer-hit counts, before/after/removed, per
# sample x primer x read file x orientation
add_sheet_to_excel(
  workbook_path = output_excel_path,
  sheet_name = "Primer_Hit_Counts",
  data = primer_hit_counts_dt,
  rownames = FALSE,
  overwrite = TRUE
)
```

```{r echo=FALSE}
cat("Added sheet: Primer_Hit_Counts\n")
```

```{r echo=FALSE}
cat("\nResults exported to:\n",
    "File:", output_excel_path, "\n")
```

## Document and Export Column Dictionary {#column-dictionary}

Build a trailing `Column_Dictionary` sheet documenting every column of every sheet just written to `output_excel_path`, so the workbook explains itself to collaborators.

```{r build-column-dictionary, include=FALSE}
# Sheet 1: Primer_Hit_Counts
primer_hit_counts_descriptions <- c(
  Biological_Relevance = "Whether this row's primer / read-file / orientation combination has a real biological mechanism ('Biologically relevant' = forward or reverse primer at the 5' start of its own read, or read-through of the opposite primer's reverse complement at the 3' end) or none ('Contaminant / other' = orientation/read-file combinations with no expected mechanism, indicating contamination or artefact).",
  SampleID = "Sample identifier extracted from the raw FASTQ filename by extract_sample_id().",
  Primer = "Which configured primer this row counts: FWD (forward_primer) or REV (reverse_primer).",
  Read_File = "Which read file was searched: R1 (forward reads) or R2 (reverse reads).",
  Orientation = "Which of the four sequence orientations (Forward, Complement, Reverse, ReverseComplement) of the primer was searched for.",
  Hits_Before = "Number of RAW reads with this primer orientation found within primer_anchor_buffer bases of a read end (anchored, exact IUPAC match), before cutadapt trimming.",
  Hits_After = "Same anchored count on the cutadapt-trimmed reads; for the biologically meaningful values this should be approximately 0 after successful trimming.",
  Hits_Removed = "Hits_Before minus Hits_After: primer-bearing reads cleaned for this sample/primer/read-file/orientation combination.",
  Pct_Removed = "Percentage of primer-bearing reads removed relative to the pre-trimming count (100 * Hits_Removed / Hits_Before, rounded to 2 decimal places); NA when Hits_Before is 0 (nothing to remove)."
)

primer_hit_counts_dict <- build_column_dictionary(
  sheet_name = "Primer_Hit_Counts",
  data = primer_hit_counts_dt,
  descriptions = primer_hit_counts_descriptions,
  workbook_path = output_excel_path,
  rownames = FALSE
)

# Combine the per-sheet dictionaries (only one sheet has data columns here)
column_dictionary <- primer_hit_counts_dict
```

```{r export-column-dictionary, include=FALSE}
# Sheet 2: Column_Dictionary (written last, so it documents all sheets above)
add_sheet_to_excel(
  workbook_path = output_excel_path,
  sheet_name = "Column_Dictionary",
  data = column_dictionary,
  rownames = FALSE,
  overwrite = TRUE
)
cat("Added sheet: Column_Dictionary\n")
```

```{r output-links-excel, echo=FALSE}
render_output_links(output_excel_path, labels = "Primer trimming summary (Excel workbook)")
```

------------------------------------------------------------------------

# Output File Summary

The tree below lists every file this notebook has written to its own output folder, [results/3_cutadapt_primer_trimming/](../../results/3_cutadapt_primer_trimming/), as a clickable, portable link (relative to this notebook's own location) with a short description -- built live from what is actually on disk at knit time via the project's shared [render_output_tree_function.R](../functions/render_output_tree_function.R) helper, so it always matches this run's real output rather than a hand-maintained list.

```{r output-summary, echo=FALSE}
render_output_tree(
  output_folder,
  descriptions = c(
    "primer_trimmed_reads" = "Primer-trimmed FASTQ reads (folder, input to Steps 4 & 5)",
    "cutadapt_logs" = "Cutadapt log files (folder)",
    "primer_trimming_summary.xlsx" = "Primer trimming summary (before/after primer-hit counts, Excel workbook)"
  )
)
```

------------------------------------------------------------------------

# Recommended Next Step

The primer-trimmed FASTQ files and per-sample primer-hit diagnostics exported above are ready for the next stages. First use the [Step 4 — DADA2 Parameter Explorer](4_dada2_parameter_selection.md) to visualize the amplified target, inspect read quality and retention, validate representative samples if needed, and save filtering parameters. Then proceed to [Step 5 — DADA2 Pipeline](5_dada2_pipeline.md), which reads this notebook's [primer_trimmed_reads/](../../results/3_cutadapt_primer_trimming/primer_trimmed_reads) folder and performs length filtering, N-filtering, quality filtering, denoising, merging, chimera removal, ASV table construction, and taxonomy assignment.

------------------------------------------------------------------------

# Session Information

Record the R environment for reproducibility.

```{r session-info, echo=FALSE}
sessionInfo()
```

------------------------------------------------------------------------

# References

## Methods

-   Martin M (2011). Cutadapt removes adapter sequences from high-throughput sequencing reads. *EMBnet.journal*, 17(1), 10-12. <https://doi.org/10.14806/ej.17.1.200> — primary citation for the [Cutadapt](https://cutadapt.readthedocs.io/en/stable/) tool this notebook runs.
-   Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP (2016). DADA2: High-resolution sample inference from Illumina amplicon data. *Nature Methods*, 13, 581-583. <https://doi.org/10.1038/nmeth.3869> — see also the [DADA2 ITS Pipeline Workflow](https://benjjneb.github.io/dada2/ITS_workflow.html) documentation on removing primers before running DADA2.
-   Pagès H, Aboyoun P, Gentleman R, DebRoy S. [Biostrings](https://bioconductor.org/packages/release/bioc/html/Biostrings.html): Efficient manipulation of biological strings. Bioconductor. — provides `vcountPattern()`, used by this notebook's custom anchored primer-hit counter.

## Related

-   [Step 1 — Data Integrity Check](1_data_integrity_check.md) — optional prerequisite; validates FASTQ files before this notebook runs.
-   [Step 2 — FastQC Quality Reports](2_fastqc_quality_reports.md) — optional prerequisite; read-quality inspection before trimming.
-   [Step 4 — DADA2 Parameter Explorer](4_dada2_parameter_selection.md) — optional prerequisite; visualize the amplified target, inspect read quality and retention and save filtering parameters
-   [Step 5 — DADA2 Pipeline](5_dada2_pipeline.md) — this notebook's recommended next step; consumes the primer-trimmed reads produced here.

------------------------------------------------------------------------

# Appendix: Troubleshooting Guide

## Common Issues and Solutions

### Cutadapt Not Found

**Error**: `Cutadapt executable not found or not executable` or `Cutadapt could not be started`

**Solutions**:

- Recreate the project-local installation with `Rscript setup/install_required_tools.R --force-reinstall`
- Verify that `cutadapt_path` points to the recreated project-local executable
- Do not copy `tools/cutadapt/venv/` between project locations or machines; Python virtual environments contain absolute interpreter paths

### Primers Still Present After Trimming

**Symptom**: In `Primer_Hit_Counts`, the biologically meaningful cells (see below) still show substantial `Hits_After`

**Possible causes**:

- Primer sequences in `forward_primer`/`reverse_primer` do not match the primers actually used
- `primer_anchor_buffer` is too small for a library prep with several leading bases before the primer (raising the buffer would catch primers sitting a little further from the read end)
- Read length is shorter than the primer, so the primer cannot be fully contained in the read

**Actions**:

- Read the cutadapt logs ([`cutadapt_logs/`](../../results/3_cutadapt_primer_trimming/cutadapt_logs/)): the "=== Summary ===" and per-adapter "Trimmed: N times" lines are cutadapt's own error-tolerant tally. If cutadapt reports it trimmed the primers but the anchored count is still high, the primer sequence or anchor window is the likely culprit
- Verify primer sequences are correct for the experiment's amplicon region

### Low or Zero Primer Counts Even Before Trimming

**Symptom**: `Hits_Before` is near zero in the meaningful values

**Likely cause**: Primers were already removed upstream (e.g. by the sequencing facility), so there is nothing to trim. This reflects the input data and is not a notebook malfunction. Confirm with the cutadapt logs (they will also report near-zero matches) and by searching a few raw reads directly.

### Memory Issues

**Error**: Cannot allocate memory for FASTQ processing

**Solutions**:

- Reduce `nr_threads` to use fewer parallel processes
- Reduce `fastq_chunk_size` so each worker holds fewer reads at once
- Process samples in batches

## Understanding the Output

### Which `Primer_Hit_Counts` values matter

The table reports all 32 combinations per sample (2 primers x 2 read files x 4 orientations x before/after collapsed into columns), but only four have a real biological mechanism and are expected to be non-zero *before* trimming:

- **FWD primer, R1, Forward**: the forward primer at the 5' start of forward reads (the primary event; expected high before, \~0 after)
- **REV primer, R2, Forward**: the reverse primer at the 5' start of reverse reads (the primary event; expected high before, \~0 after)
- **REV primer, R1, ReverseComplement**: reverse-primer read-through at the 3' end of forward reads (expected only when the insert is shorter than the read length; \~0 after)
- **FWD primer, R2, ReverseComplement**: forward-primer read-through at the 3' end of reverse reads (expected only for short inserts; \~0 after)

All other values (Complement/Reverse orientations, and each primer in the "wrong" read file) have no biological mechanism and should be at or near zero both before and after -- they exist as a contamination/library-prep sanity net.

### A note on exact matching

`Primer_Hit_Counts` uses an anchored, exact (zero-mismatch) IUPAC search, so a read whose primer region carries a sequencing error may not be counted; the `Hits_Before` values are therefore a conservative lower bound. The cutadapt logs ([`cutadapt_logs/`](../../results/3_cutadapt_primer_trimming/cutadapt_logs/)) hold cutadapt's own error-tolerant tally for comparison.
